Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-05 appendix.tex
re-writing sec 7 in progress

\subsection{Marginal stability for the fundamental
  mode}\label{bcrit_alt}
For the fundamental mode ($M=0$) we can deduce $\beta_c\ll 1$, which
motivates us to make this strong assumption from the onset. In this
case, Eq. \ref{relax_cond1} --- \ref{relax_cond2} 
imply the following balance for large $\khat$ and small
$\hat{\omega}_c$ while $\hat{\omega}_c\khat$ remain fixed:  
\begin{subequations}\begin{align}
    &0  = \left[1 + \gamma\beta_c^2\left(1-\gamma\right)\right] + 2h q
    \gamma \beta_c \left(\hat{\omega}_c\khat\right) - \left(\hat{\omega}_c\khat\right)^2,\\
    &0 =  \beta_c(1-\gamma) + hq \left(\hat{\omega}_c\khat\right). 
  \end{align}\end{subequations}
Then
\begin{align}
  \beta_c = \frac{h|q|}{\gamma-1}\left(1 - \frac{\gamma h^2 
      q^2}{\gamma-1}\right)^{-1/2},
\end{align}
which reduces to Eq. \ref{iso_vsi_cond} at $O(h)$.

@@

\section{Fiducial model for a protoplanetary disk}\label{mmsn}
For results application in \S\ref{application} we use the disk model
described in \cite{chiang10}. This disk model orbits a Solar-mass star and 
has the surface density distribution
\begin{align}\label{mmsn_sigma}
  \Sigma = 2200
  \hat{\Sigma}\left(\frac{r}{\mathrm{AU}}\right)^{-3/2}\mathrm{g}\,\mathrm{cm}^{-2},  
\end{align}
and the temperature profile
\begin{align}\label{mmsn_temp}
  T = 120\hat{T}\left(\frac{r}{\mathrm{AU}}\right)^{-3/7} \mathrm{K},
\end{align}
which implies $q=-3/7$. In the above expressions, $\hat{\Sigma}$ and
$\hat{T}$ are dimensionless coefficients used to scale the model
relative to the MMSN. Their nominal values are unity. By assuming a
vertically isothermal disk, we deduce the disk aspect-ratio 
\begin{align}\label{mmsn_epsilon}
  h =
  3.36\times10^{-2}\left(\frac{\hat{T}}{\mu}\right)^{1/2}\left(\frac{r}{\mathrm{AU}}\right)^{2/7}, 
\end{align}
and the mid-plane density distribution 
\begin{align}
%  \rho_0 = 2.7\times10^{-9}
%  \hat{\upsilon}\left(\frac{r}{\mathrm{AU}}\right)^{-39/14}\mathrm{g}\,\mathrm{cm}^{-3},  
\rho_0 = 1.7\times10^{-9}
  \hat{\Sigma}\left(\frac{\hat{T}}{\mu}\right)^{-1/2}\left(\frac{r}{\mathrm{AU}}\right)^{-39/14}\mathrm{g}\,\mathrm{cm}^{-3},
\end{align}
which implies $p=-39/14$. 
%0.033576258
In addition, we use the opacity model
 \begin{align}
   \kappa_d &= 2 \hat{\kappa}_d \left(\frac{T}{100\mathrm{K}}\right)^2
   \mathrm{cm}^2\,\mathrm{g}^{-1}
    =
   2.88\hat{\kappa}_d\hat{T}^2\left(\frac{r}{\mathrm{AU}}\right)^{-6/7}\mathrm{cm}^2\,\mathrm{g}^{-1},   
 \end{align}
\citep{bell94} where the second equality follows from the model
temperature profile above, and $\hat{\kappa}_d$ has similar meaning as
$\hat{\Sigma}$ and $\hat{T}$.

>>>>

% \subsection{Marginal stability for the fundamental
%   mode}\label{bcrit_alt}
% For the fundamental mode ($M=0$) we can deduce $\beta_c\ll 1$, which
% motivates us to make this strong assumption from the onset. In this
% case, Eq. \ref{relax_cond1} --- \ref{relax_cond2} 
% imply the following balance for large $\khat$ and small
% $\hat{\omega}_c$ while $\hat{\omega}_c\khat$ remain fixed:  
% \begin{subequations}\begin{align}
%     &0  = \left[1 + \gamma\beta_c^2\left(1-\gamma\right)\right] + 2h q
%     \gamma \beta_c \left(\hat{\omega}_c\khat\right) - \left(\hat{\omega}_c\khat\right)^2,\\
%     &0 =  \beta_c(1-\gamma) + hq \left(\hat{\omega}_c\khat\right). 
%   \end{align}\end{subequations}
% Then
% \begin{align}
%   \beta_c = \frac{h|q|}{\gamma-1}\left(1 - \frac{\gamma h^2 
%       q^2}{\gamma-1}\right)^{-1/2},
% \end{align}
% which reduces to Eq. \ref{iso_vsi_cond} at $O(h)$.

@@

\section{Fiducial model for a protoplanetary disk}\label{mmsn}
For results application in \S\ref{application} we use the disk model
described in \cite{chiang10}. This disk model orbits a Solar-mass star and 
has the surface density distribution
\begin{align}\label{mmsn_sigma}
  \Sigma = 2200
  \hat{\Sigma}\left(\frac{r}{\mathrm{AU}}\right)^{-3/2}\mathrm{g}\,\mathrm{cm}^{-2},  
\end{align}
and the temperature profile
\begin{align}\label{mmsn_temp}
  T = 120\hat{T}\left(\frac{r}{\mathrm{AU}}\right)^{-3/7} \mathrm{K},
\end{align}
which implies $q=-3/7$. In the above expressions, $\hat{\Sigma}$ and
$\hat{T}$ are dimensionless coefficients used to scale the model
relative to the MMSN. By assuming a
vertically isothermal disk, we deduce the disk aspect-ratio 
\begin{align}\label{mmsn_epsilon}
  h =
  3.36\times10^{-2}\left(\frac{\hat{T}}{\mu}\right)^{1/2}\left(\frac{r}{\mathrm{AU}}\right)^{2/7}, 
\end{align}
and the mid-plane density distribution 
\begin{align}
%  \rho_0 = 2.7\times10^{-9}
%  \hat{\upsilon}\left(\frac{r}{\mathrm{AU}}\right)^{-39/14}\mathrm{g}\,\mathrm{cm}^{-3},  
\rho_0 = 1.7\times10^{-9}
  \hat{\Sigma}\left(\frac{\hat{T}}{\mu}\right)^{-1/2}\left(\frac{r}{\mathrm{AU}}\right)^{-39/14}\mathrm{g}\,\mathrm{cm}^{-3},
\end{align}
which implies $p=-39/14$. 
%0.033576258
In addition, we use the opacity model
 \begin{align}
   \kappa_d &= 2 \hat{\kappa}_d \left(\frac{T}{100\mathrm{K}}\right)^2
   \mathrm{cm}^2\,\mathrm{g}^{-1}
    =
   2.88\hat{\kappa}_d\hat{T}^2\left(\frac{r}{\mathrm{AU}}\right)^{-6/7}\mathrm{cm}^2\,\mathrm{g}^{-1},   
 \end{align}
\citep{bell94} where the second equality follows from the model
temperature profile above, and $\hat{\kappa}_d$ has similar meaning as
$\hat{\Sigma}$ and $\hat{T}$. In \S\ref{application} we fix
$\hat{\Sigma} = \hat{T}=1$, $\mu=2.33$ and use $\gamma=1.4$.

### 2015-05-05 application.tex
re-writing sec 7 in progress

\section{Application to protoplanetary disks}\label{application} 
% \emph{I suggest the following edits}
% \emph{For this section. There are 3 main ideas, and I think they should be the subsections, more comments related to this are in the text itself}:
% \begin{itemize}
% \item Developing the cooling model in opt thin and this regimes, including MMSN value
% \item Comparison to expected $\beta_{crit}$
% \item Linear growth with MMSN cooling law
% \end{itemize}

% \emph{Suggested edits for this intro}
% \begin{itemize}
% \item Instead of starting over again with the full energy equation focus on giving a phyiscal values for $\Lambda$ and $t_c$ already introduced.
% \item Don't describe what we don't do, especially at the outset.  Say what's done, and then if essential list caveats.
% \end{itemize}

@@

% \emph{I think 7.1-7.4 should be combined to a single section on the "Cooling Model".  As that clearly is a single idea and osme of the subsections are too trivial. Conversion to subsubsections as needed. However the application to beta-crit could be split off as 7.2.  Otherwise that material which has a crucial figure discussion is a bit buried.}

@@

\subsubsection{A model for combined thermal relaxation}\label{toy_relax}
%To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, 
We define the effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
  % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$. Writing the
volume density $\rho = \Sigma\hat{\rho}/\sqrt{2\pi}H$, where
$\hat{\rho} = \rho/\rho_0$ and $\Sigma(r)$ is the surface density at the
radius of interest, the dimensionless thermal
relaxation time $\beta$ becomes 
\begin{align}\label{real_beta}
  \beta(r,z;\khat) \equiv t_c\OmK =
  % \frac{1}{\hat{\eta}}\left[\frac{l_\mathrm{rad}^2}{3H^2}
  %   + \frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right]  \simeq
  \frac{\Sigma^2\OmK}{\eta\rho^2}\left[\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{\hat{\rho}^2(z)}{2\pi \khat^2}\right].
\end{align}
Note that $\beta$ depends the perturbation lengthscale $\khat$,
distance from the midplane, as well as the distance from the star.

@@

\subsection{Comparing $\beta$ to $\beta_\mathrm{crit}$ in the MMSN} 
Here we compare the above thermal timescale, Eq. \ref{beta_mmsn_simp}, to  
the VSI criterion Eq. \ref{iso_vsi_cond} evaluated for the MMSN,     
\begin{align}\label{bcrit_mmsn}
  % \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
  % -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
  \beta_\mathrm{crit,MMSN} = 0.024r_\mathrm{AU}^{2/7}. 
\end{align}

% \emph{This paragraph currently merely copies the figure caption, which is a waste and not illuminating to read.  Better to give some description of what the figure is showing. (see below for ideas)}\\
% In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn}  with $\mu
% =2.33$ $\gamma=1.4$, $\hat{T}=\hat{\Sigma}=1$ for opacity scales
% $\hat{\kappa}_d=0.1, \,1 $ (MMSN) and $10$, at $z=0$ and $z=3H$. \\
% \emph{Here you can describe the general features e.g. diffusive and thin cooling regions, which will help later discussions.}\\
% We also plot Eq. \ref{bcrit_mmsn} in each panel.  \emph{This sentence explains nothing and is likely to be of little help to the reader.  Instead wait to describe the science, e.g. after describing the general shape of the $\beta$ curves,  you can start the discussion below with "Fig () also compares the MMSN cooling times to $\beta_{crit}$.}

Consider first the MMSN shown in the middle panels of 
Fig. \ref{mmsn_bcrit_bcool}. The critical thermal timescale,
Eq. \ref{bcrit_mmsn}, is useful in the optically thin regime (where the
curves overlap), since in this limit $\beta$ does not depend on
height, as assumed in our analysis. \\
\emph{above sentence too long and confusing.  "useful" is a vague term.  
"where the curves overlap" is also confusing, forcing the reader to look at the figure in the middle of a complex sentence.  Better to have already explained above how the $\beta$ curves show that opt. thin cooling is height indep. (even though should be obvious from equations).  Then the comparison to $\beta_{crit}$ can be made as something like "Since beta-crit was derived for a height-independent cooling time,  the direct comparison our MMSN cooling model is valid only for modes with optically thin cooling. "} 
We thus infer that the VSI can
operate at tens of AU in the MMSN.

In the inner few AU of the MMSN, however, $\beta$ 
may be smaller or greater than $\beta_\mathrm{crit}$ depending on
$z$ as the disk transitions from an optically thick midplane to an
optically thin atmosphere (see also Fig. \ref{bcrit_mmsn2d}). 
%Given a fixed $\khat$, one can have $\beta > \beta_\mathrm{crit}$ at
%the optically thick midplane, but with increasing height the disk 
%becomes optically thin and $\beta<\beta_\mathrm{crit}$. 
Our analytical criteria $\beta < \beta_\mathrm{crit}$ is not valid for
vertically non-uniform $\beta$, and we must turn to numerical
solutions of the linear problem to determine whether or not the VSI
operates. We do this in the next section.    \\
\emph{a better organized discussion might end on this point.  Might not be easy to rearrange to do this, but it would ideal to have the seamless flow to what you do next.}

Lowering the opacity, as shown in the top panels of
Fig. \ref{mmsn_bcrit_bcool}, shifts the curves inwards but also raises
them.\\
\emph{this graphical discussion is not very inuitive (the main problem) and even arbitrary (lowering vs. shifting to the left).  Better to say what's happening physically, i.e. `With higher opacity, optically thin cooling faster and diffusive cooling is slower, explaining how the curves shift.} \\
In fact, lowering the opacity by an order of magnitude renders  
$\beta > \beta_\mathrm{crit}$ even away from the midplane. This is 
expected to stabilize the disk. Raising the opacity, as shown in 
the lower panels of Fig. \ref{mmsn_bcrit_bcool}, decreases the thermal
timescales, but shifts the curves outward, so the instability is
restricted to larger radii.

\begin{figure*}
  \includegraphics[scale=.47,clip=true,trim=0cm 1.8cm 0cm
  0cm]{figures/bcrit_mmsn_kap0d1_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm 1.8cm 0cm
  0cm]{figures/bcrit_mmsn_kap0d1_z3}\\ 
  \includegraphics[scale=.47,clip=true,trim=0cm 1.8cm 0cm
  1cm]{figures/bcrit_mmsn_kap1_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm
  1.8cm 0cm 1cm]{figures/bcrit_mmsn_kap1_z3}\\
  \includegraphics[scale=.47,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/bcrit_mmsn_kap10_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm 0cm 0cm
  1cm]{figures/bcrit_mmsn_kap10_z3} 
  \caption{Dimensionless thermal relaxation timescales $\beta$,
    evaluated at the midplane (left) and at $z=3H$ (right) in the
    fiducial PPD. Eq. \ref{beta_mmsn} is plotted  
    for three values of the 
    perturbation radial wavenumber: $\khat=1$ (dotted), $\khat=10$
    (dashed) and $\khat=100$ (dashed-dot), for three values of the
    opacity relative to the MMSN: $\hat{\kappa}_d=0.1$ (top),
    $\hat{\kappa}_d=1$ (middle) and $\hat{\kappa}_d=10$ (bottom).  
    The solid line is the 
    critical thermal relaxation timescale $\beta_\mathrm{crit}$.  
    \label{mmsn_bcrit_bcool}}   
\end{figure*}

\emph{the following statement is not generally true, so it should not be stated without caveat}

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$ 
for sufficiently large $\khat$. We illustrate this in 
Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at each radius the value
of $\khat$ such that $\beta = \beta_\mathrm{crit}$. If we consider the
midplane of the inner disk, say $r\sim 1$AU, the VSI would only occur
on very small radial lengthscales $\ll O(10^{-2}H)$. In the disk
atmosphere where $\beta$ is small, surface modes may develop, 
%However, at such
%small scales, surface modes are likely to dominate (and are
%restricted to the disk atmosphere), 
but they are not well-defined in vertically isothermal disk models
with an imposed boundary, and they also require large $\khat$.

%surface modes can develop for large $\khat$, 
%which have small vertical lengthscales scales. 
%also small vertical scales if they are surface modes

\emph{The following viscosity discussion might be better placed elsewhere.  Since the issue also comes up for numerical solutions (not just the comparison to beta-crit here), maybe a short concluding subsection to summarize all the arguments on the "vialibity of short wavelength VSI".}

However, disturbances with large $\khat$ may be subject to viscous
decay in a real disk. Adopting the standard precription for the
kinematic viscosity $\nu = \alpha c_s H$ \citep{shakura73}, where
$\alpha$ is the dimensionless viscosity parameter, 
the viscous timescale for a perturbation lengthscale 
$l\sim 1/k_x$ is $t_\mathrm{visc} = 1/\alpha\khat^2\Omega $. 
Setting $t_\mathrm{visc}$ to be longer than the
minimum VSI growth timescale $1/h\Omega$ implies 
$\khat^2\lesssim h/\alpha$. Inserting $h \sim 0.05$,
we find only modes with $\khat\lesssim 20$ are relevant for
$\alpha\sim 10^{-4}$ or $\khat\lesssim 70$ for $\alpha\sim
10^{-5}$. \citepalias[These low viscosity levels are required for the
VSI, see][]{nelson13}.

@@

\subsection{VSI in the MMSN}
Here we solve the linear stability problem in the MMSN with the scale and
height dependent thermal relaxation timescale given by
Eq. \ref{beta_mmsn_simp}. 
 Fig. \ref{beta_compare} show examples of
the thermal timescales at $r=5$AU and $r=50$AU for a
perturbation wavenumber $\khat=30$. We also plot the 
corresponding critical thermal timescales evaluated for the MMSN
(Eq. \ref{bcrit_mmsn}).  \\
\emph{For this figure to belong here (and not in the previous subsection, with the other comparisons of $\beta$ and $\beta_{crit}$) the discussion should be be more focused on the linear analysis.  Again it's better to start by discussing what the figure tells us, than repeating the figure caption. E.g.\ you could use Fig 21 to explain when a height varying $\beta$ is required.  I would also explain (especially if it stays in this subsection) what the growth time is for the r = 5 AU case.  Looks to be $\sim 1000$ orbits, which is remakrable given that $\beta < \beta_{\rm crit}$ outside $z \gtrsim  1.6 H$.?  That kind of detail could be illuminating, as it's nominally surprising given previous evidence that behavior near 5 -- 7 $H$ can affect growth.}

The outer disk ($r=50$AU) is optically thin,
$\beta<\beta_\mathrm{crit}$  and  does not depend on height.
Accordingly we expect, and find, the VSI operates at $r=50$AU. The
inner disk ($r=5$AU) is optically thick near the midplane but becomes
optically thin away from it, such that $\beta < \beta_\mathrm{crit}$
for $|z|\gtrsim1.25H$. In fact, we also find the VSI in this case, but
with much reduced growth rates (when scaled by the local rotation), as
described below.

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\sigma$ of the fundamental mode with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\OmK$. In fact, we either only find the fundamental mode, or 
that it is the most unstable, except for the appearance of some 
surface modes for large wavenumbers $\khat\gtrsim 50$, which are 
boundary-dependent.  \\
\emph{above sentence too long and complex, could e.g.\ split off that surface modes are neglected and why.}
%However, surface modes are dependent on
%boundary conditions in a vertically isothermal disk \citep{barker15},  
%and 
Disturbances with large $\khat$ are also more likely subject to viscous 
decay, as discussed above.

% It is also unclear if surface modes are
% dynamically important given they are concentrated in the disk
% atmosphere where the density is small. 
% The dominance of the fundamental mode is also
% observed in hydrodynamic simulations of \cite{stoll14} which treat
% radiative cooling properly.   unclear if surface modes are
% dynamically important because the there is little mass in the disk at

We therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in outer parts of the MMSN. We see the VSI operates from $\sim 5$AU to $\sim
50$AU with growth timescales $\sim 30$---40 orbits and
$t_\mathrm{grow}$ rises rapidly towards the inner disk. It is
necessary to consider smaller scales with decreasing $r$ as the disk
generally becomes optically thick.

The fact that we obtain the VSI in the inner disk despite $\beta >
\beta_\mathrm{crit}$ near the midplane suggests that the instability
can operate provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
become much longer. For example, for $\khat=30$ we find
$t_\mathrm{grow}\sim 40$ orbits at $r=50$AU, but $t_\mathrm{grow}\sim
400$ orbits at $r=5$AU. Thus, introducing an optically thick midplane
significantly reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane.

>>>>

\section{Application to protoplanetary disks}\label{application}

@@

@@

\subsubsection{Evaluation for a fiducial disk model}\label{toy_relax}
We define the effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
  % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$. Writing the
volume density $\rho = \Sigma\hat{\rho}/\sqrt{2\pi}H$, where
$\hat{\rho} = \rho/\rho_0$ and $\Sigma(r)$ is the surface density at the
radius of interest, the dimensionless thermal
relaxation time $\beta$ becomes 
\begin{align}\label{real_beta}
  \beta(r,z;\khat) \equiv t_c\OmK =
  % \frac{1}{\hat{\eta}}\left[\frac{l_\mathrm{rad}^2}{3H^2}
  %   + \frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right]  \simeq
  \frac{\Sigma^2\OmK}{\eta\rho^2}\left[\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{\hat{\rho}^2(z)}{2\pi \khat^2}\right].
\end{align}
Note that $\beta$ depends the perturbation lengthscale, 
distance from the midplane, as well as the distance from the star.

@@

\subsection{Comparing $\beta$ to $\beta_\mathrm{crit}$ in the MMSN}

Fig. \ref{beta_compare} show examples of
$\beta$ at $r=5$AU and $r=50$AU for a perturbation wavenumber
$\khat=30$.

% In the outer disk  at $r=50$AU, $\beta$ is vertically-uniform and is less
% than $\beta_\mathrm{crit}$.  Accordingly we expect, and find, the VSI
% operates.    
% In the inner disk ($r=5$AU), the perturbation is in the optically
% thick regime near the midplane, but becomes optically-thin away from
% the midplane.  In this case, $\beta < \beta_\mathrm{crit}$
% for $|z|\gtrsim1.25H$. In fact, we also find the VSI in this case, but
% with much reduced growth rates (when scaled by the local rotation), as
% described below.

In Fig. \ref{mmsn_bcrit_bcool} we compare the above thermal timescale,
Eq. \ref{beta_mmsn_simp},  for several opacity scales, to  
the VSI criterion Eq. \ref{iso_vsi_cond} evaluated for the MMSN,     
\begin{align}\label{bcrit_mmsn}
  % \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
  % -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
  \beta_\mathrm{crit} = 0.024r_\mathrm{AU}^{2/7}. 
\end{align}

\begin{figure*}
  \includegraphics[scale=.47,clip=true,trim=0cm 1.8cm 0cm
  0cm]{figures/bcrit_mmsn_kap0d1_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm 1.8cm 0cm
  0cm]{figures/bcrit_mmsn_kap0d1_z3}\\ 
  \includegraphics[scale=.47,clip=true,trim=0cm 1.8cm 0cm
  1cm]{figures/bcrit_mmsn_kap1_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm
  1.8cm 0cm 1cm]{figures/bcrit_mmsn_kap1_z3}\\
  \includegraphics[scale=.47,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/bcrit_mmsn_kap10_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm 0cm 0cm
  1cm]{figures/bcrit_mmsn_kap10_z3} 
  \caption{Dimensionless thermal relaxation timescales $\beta$,
    evaluated at the midplane (left) and at $z=3H$ (right) in the
    fiducial PPD. Eq. \ref{beta_mmsn_simp} is plotted  
    for three values of the 
    perturbation radial wavenumber: $\khat=1$ (dotted), $\khat=10$
    (dashed) and $\khat=100$ (dashed-dot), for three values of the
    opacity relative to the MMSN: $\hat{\kappa}_d=0.1$ (top),
    $\hat{\kappa}_d=1$ (middle) and $\hat{\kappa}_d=10$ (bottom).  
    The solid line is the 
    critical thermal relaxation timescale $\beta_\mathrm{crit}$.  
    \label{mmsn_bcrit_bcool}}   
\end{figure*}

For disturbances in the optically-thin cooling regime, $\beta$ does
not depend on the height $z$ or perturbation lengthscale $\khat$. This
generally occurs at sufficiently large radii. In this case we may
directly compare $\beta$ to $\beta_\mathrm{crit}$, which was derived
assuming it is vertically-uniform.  For the case with MMSN
opacity ($\hat{\kappa}_d=1$), we  infer that the VSI can operate at
10s of AU.

For disturbances in the optically-thick or diffusive regime, $\beta$
increases with decreasing $\khat$ (larger scales) and decreases with
increasing height, on account of the density-dependence of
$t_\mathrm{diff}$.  This generally occurs at small radii. In this case 
$\beta$ may be smaller or greater than $\beta_\mathrm{crit}$ depending
on $z$. Our analytical criteria $\beta < \beta_\mathrm{crit}$ is not valid for
vertically non-uniform $\beta$, and we must turn to numerical
solutions of the linear problem to determine whether or not the VSI
operates. We do this in \S\ref{vsi_mmsn_grow}.

%\emph{a better organized discussion might end on this point.  Might not be easy to rearrange to do this, but it would ideal to have the seamless flow to what you do next.}

Next consider the effect of the opacity scale $\hat{\kappa}_d$. The previous section
(see also Eq. \ref{beta_mmsn_simp}) indicates that increasing $\hat{\kappa}_d$
increases (decreases) the cooling time in the optically thick (thin)
limits. The overall effect of  increasing (decreasing) the opacity
scale by an order of magnitude is illustrated in the lower (upper)
panels of Fig. \ref{mmsn_bcrit_bcool}.

For increasing opacity, $\hat{\kappa}_d=1\to 10$,  a comparison
between the middle and lower panels shows that the optically-thick
regions shift outward (where the three curves are distinct); while the
optically-thin regions have lowered cooling times (where the curves
overlap). Thus, instability is expected to be restricted to larger
radii.

For decreasing opacity, $\hat{\kappa}_d=1\to 0.1$,  the
optically-thick regions shift inwards; while cooling times in the
optically-thin limit are increased. In fact, the latter effect is
significant enough to raise $\beta > \beta_\mathrm{crit}$ over all
radii, and even away from the midplane. This is expected to stabilize
the disk against the VSI.

%\emph{the following statement is not generally true, so it should not be stated without caveat}

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
at small radii for sufficiently large $\khat$. We illustrate this in 
Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at each radius the value
of $\khat$ such that $\beta = \beta_\mathrm{crit}$. If we consider the
midplane at, say $r\sim 1$AU, the VSI would only occur on very small
radial lengthscales $\ll O(10^{-2}H)$. It is, however, unclear if such
small-scale disturbances are dynamically important.

%In the disk
% atmosphere where $\beta$ is small, surface modes may develop, 
% but they are not well-defined in vertically isothermal disk models
% with an imposed boundary, and they also require large $\khat$.

@@

\subsection{VSI growth in the MMSN}\label{vsi_mmsn_grow}
Here we solve the linear stability problem in the MMSN with the scale and
height dependent thermal relaxation timescale given by
Eq. \ref{beta_mmsn_simp}.

% \\
% \emph{For this figure to belong here (and not in the previous subsection, with the other comparisons of $\beta$ and $\beta_{crit}$) the discussion should be be more focused on the linear analysis.  Again it's better to start by discussing what the figure tells us, than repeating the figure caption. E.g.\ you could use Fig 21 to explain when a height varying $\beta$ is required.  I would also explain (especially if it stays in this subsection) what the growth time is for the r = 5 AU case.  Looks to be $\sim 1000$ orbits, which is remakrable given that $\beta < \beta_{\rm crit}$ outside $z \gtrsim  1.6 H$.?  That kind of detail could be illuminating, as it's nominally surprising given previous evidence that behavior near 5 -- 7 $H$ can affect growth.}

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\sigma$ of the fundamental mode with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\OmK$. In fact, we either only find the fundamental mode, or 
that it is the most unstable, except for the appearance of some 
surface modes for large wavenumbers $\khat\gtrsim 50$. We do not
consider surface modes since they are boundary-dependent and likely
subject to viscous decay (see). 
%  which are 
% boundary-dependent. 
% Disturbances with large $\khat$ are also more likely subject to viscous 
% decay, as discussed above. 
 %\\
%\emph{above sentence too long and complex, could e.g.\ split off that surface modes are neglected and why.}
%However, surface modes are dependent on
%boundary conditions in a vertically isothermal disk \citep{barker15},  
%and

We therefore consider Fig. \ref{mmsn_overall} as a representation of the
VSI in the fiducial MMSN. We see the VSI operates from $\sim 5$AU to $\sim
50$AU with growth timescales $\sim 30$---40 orbits and
$t_\mathrm{grow}$ rises rapidly towards the inner disk. It is
necessary to consider smaller scales with decreasing $r$ as
perturbations generally become optically thick.

The fact that we obtain the VSI in the inner disk despite $\beta >
\beta_\mathrm{crit}$ near the midplane suggests that the instability
can operate provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
become much longer. For example, for $\khat=30$ we find
$t_\mathrm{grow}\sim 40$ orbits at $r=50$AU, but $t_\mathrm{grow}\sim
400$ orbits at $r=5$AU (and rapidly increases to $> 10^3P_0$ for
smaller $r$). Thus, introducing an optically thick midplane
significantly reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane.
